### 2014-02-22 reanalysis/ComBat_original/combat_paper_dataset2_reproduced.rmd

This report aims to show to what extent the use of ComBat led to false results in the second "proof of concept" analysis performed in Johnson et al. The example named "Data Set 2" and their analysis are described in the supplementary materials for Johnson et al.
Description on how to obtain the data were found [here](https://groups.google.com/forum/#!msg/combat-user-forum/S9vBcXw8RGk/QIoD6oRBM9IJ) which had links to [dataExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/dataExample2.txt)
and [sampleInfoExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/sampleInfoExample2.txt)

@@

```{r, results='hide'}
library(pheatmap)
library(RColorBrewer)
library(sva)
library(qvalue)
library(limma)
```
```{r, cache=TRUE}

orgdatamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))
datamatrix=orgdatamatrix[1:3000,]# for dev.
#datamatrix=orgdatamatrix# for dev.
sampleannotation = read.table("data/sampleInfoExample2.txt", 
                              sep="\t", header=TRUE, stringsAsFactors=FALSE)
rownames(sampleannotation)=sampleannotation$ArrayName
sampleannotation$Batch=factor(as.character(sampleannotation$Batch)) 
sampleannotation$Cell=factor(as.character(sampleannotation$Cell)) 
# must be discrete for the pheatmap
#table(sampleannotation$ArrayName==dimnames(datamatrix)[[2]]) # ordercheck
datamatrix=datamatrix[,sampleannotation$Type!="WT"]
sampleannotation=sampleannotation[sampleannotation$Type!="WT",]
sampleannotation$Type=factor(sampleannotation$Type)
print(dim(datamatrix))
#dev/debug
useparprior=TRUE
```
The covariate/batch balance;
```{r}
print(table(sampleannotation[ , c("Batch", "Type")])  )
```

@@

```{r, tidy=FALSE}
variatonmeasure = apply(datamatrix, 1, FUN= function(x){var(x)})
clustermatrix = datamatrix[order(variatonmeasure, decreasing=TRUE),][1:2698,]
Batchcol = brewer.pal(8,"Set2")[1:3]
names(Batchcol) = levels(sampleannotation$Batch)
Typecol =brewer.pal(8,"Set2")[4:5]
names(Typecol) = levels(sampleannotation$Type)
ann_colors = list(Batch = Batchcol, Type = Typecol)
```
```{r reproduced_figA1, dev='png', fig.width=8, fig.height=15, fig.cap="fig. 1"}
pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
         color = colorRampPalette(c("red", "black", "green"))(n = 299),
         annotation = sampleannotation[, c("Batch", "Type")],  annotation_colors = ann_colors,
         cellheight=NA, cellwidth=13,fontsize=12,  drop_levels=TRUE, show_rownames=F,
         main=paste("Reproduced Fig A1.",sep=""),border_color=NA,treeheight_row=0)

```
The Heatmap does not look exactly as Fig. A1 in the paper. This could be due to undocumented transformations preformed in Johnson et al., for example log-transformation or standarization. And there are different ways of calculating variation for a probe. What parameters that were used for the clustring could also be different. But the purpose of the figure was to show that a batch effect is present in the data and espeically for batch 3. This is also achived in the reproduced figure.

@@

Batch3 = datamatrix[,sampleannotation$Batch=="3"]
```

@@

There is no description in how the 770 genes were selected so we use the same approach as in the reprodution of Figure A.1.
```{r reproduced_figA2, dev='png', fig.width=8, fig.height=15}
clustermatrix = EB3
variatonmeasure = apply(clustermatrix, 1, FUN= function(x){var(x)})
clustermatrix = clustermatrix[order(variatonmeasure, decreasing=TRUE),][1:770,]
Batchcol = brewer.pal(8,"Set2")[1:3]
names(Batchcol) = levels(sampleannotation$Batch)
Typecol =brewer.pal(8,"Set2")[4:6]
names(Typecol) = levels(sampleannotation$Type)
Cellcol =brewer.pal(8,"Set1")[1:5]
names(Cellcol) = levels(sampleannotation$Cell)
ann_colors = list(Batch = Batchcol, Type = Typecol, Cell=Cellcol)
pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
         color = colorRampPalette(c("red", "black", "green"))(n = 299),
         annotation = sampleannotation[, c("Batch", "Type", "Cell")],  
         cellheight=NA, cellwidth=13,fontsize=12,  drop_levels=TRUE, show_rownames=F,
         main=paste("Reproduced Fig A2.",sep=""),border_color=NA,treeheight_row=0)

```
Again the heatmap is not exactly as in Johnson et al, but the batch clustring is broken and the cell clones with the same number seem to cluster.

@@

We start with the data before ComBat adjustment. Limma works best with log transformed and between array normalized values, so first we have to do something with the negative signals.  Then the test is run with the batch included as a blocking factor.
```{r}
thismat = datamatrix
thismat[thismat<1]=1
thismat = log2(thismat)
thismat=normalizeBetweenArrays(thismat)
Type = as.factor(sampleannotation$Type)
Block = as.factor(sampleannotation$Batch)
design <- model.matrix(~0+Type+Block)
fit <- lmFit(thismat, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)	
fit2 = contrasts.fit(fit, cont.matrix)
datamatrix_limma_res <- eBayes(fit2)
print(table(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05))
```

@@

Since limma was not the tool used in Johnson et al.s workflow, we will first create the corresponding analysis in Limma for the whole ComBat adjusted data set in order to make a fair comparison.
```{r}
thismat = EB3
thismat[thismat<1]=1
thismat = log2(thismat)
thismat=normalizeBetweenArrays(thismat)
Type = as.factor(sampleannotation$Type)
design = model.matrix(~0 + Type)
fit <- lmFit(thismat, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
EB3_limma_res <- eBayes(fit2)
print(table(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05))
```
The number of genes below the significance threshold is fairly similar to the t-test.

@@

Lets assume one aim of this study was to obtain a list of genes that differed between control and treatment with a significant cut-off of q<0.05. Then handling the batch effect with ComBat adjustment will yield a list of `r sum(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05)` wheras a list of `r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)` probes would have been closer to the truth. Also it is of interest to note that it is mostly the same probes that are found.
```{r, tidy=FALSE}
table( qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05 ,
         qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05,
       dnn=c("limma", "ComBat"))
```
And if they were to use the top 1000 probes in a gene set test they found have used many of the same genes regardless of the use of ComBat.
```{r, tidy=FALSE}
table( rank(datamatrix_limma_res$p.value[,1])  <=1000 & 
         rank(EB3_limma_res$p.value[,1])<=1000   )
```
Our conclusion is that for this study the error introduced by the use of ComBat would only have a modest effect on the final result.

Additional sanity checks
---------
For this data set ComBat's handling of batch-effects yields only modestly different results compared to the alternative. Explanation for the difference could be that ComBat is better. To substantiate that the result from the use of ComBat is less trustworthy we provide a few additional sanity checks.

@@

fit <- lmFit(EBrand, design)
fit2 = contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
EBrand_limma_pvalues = fit2$p.value[,1]

fit <- lmFit(randommatrix, design)
fit2 = contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
randommatrix_limma_pvalues = fit2$p.value[,1]

Block = as.factor(sampleannotation$Batch)
design <- model.matrix(~0+Type+Block)
fit <- lmFit(randommatrix, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design) 
fit2 = contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)
randommatrix_limma_batch_pvalues = fit2$p.value[,1]

```
Plotting the p-values
```{r, dev='svg', fig.width=8, fig.height=8}
hist(EBrand_limma_pvalues, border="blue",  main="P-values, Random numbers", breaks=100, xlab="p-value")
hist(randommatrix_limma_pvalues, border="black", add=T, breaks=100)
hist(randommatrix_limma_batch_pvalues, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted", "No adjustment"), text.col=c("blue", "black", "red"))
```
The p-values for ComBat adjusted random numbers are enriched slighlty for low p-values. Indicating that the same enrichment seen for the real data is also false.

@@

Done!

>>>>

This <b>report</b> aims to show to what extent the use of ComBat led to false results in the second "proof of concept" analysis performed in Johnson et al. The example named "Data Set 2" and their analysis are described in the supplementary materials for Johnson et al.
Description on how to obtain the data were found [here](https://groups.google.com/forum/#!msg/combat-user-forum/S9vBcXw8RGk/QIoD6oRBM9IJ) which had links to [dataExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/dataExample2.txt)
and [sampleInfoExample2.txt](http://www.bu.edu/jlab/wp-assets/ComBat/data/sampleInfoExample2.txt)

@@

```{r, results='hide'}
includelibs = c("pheatmap", "sva", "qvalue", "limma")
lapply(includelibs, require, character.only=T)
```
```{r, cache=TRUE}

orgdatamatrix = as.matrix(read.table("data/dataExample2.txt", sep="\t", header=TRUE))
datamatrix=orgdatamatrix[1:3000,]# for dev.
sampleannotation = read.table("data/sampleInfoExample2.txt", 
                              sep="\t", header=TRUE, stringsAsFactors=FALSE)
rownames(sampleannotation)=sampleannotation$ArrayName
sampleannotation$Batch=factor(as.character(sampleannotation$Batch)) 
sampleannotation$Cell=factor(as.character(sampleannotation$Cell)) 
# must be discrete for the pheatmap
#table(sampleannotation$ArrayName==dimnames(datamatrix)[[2]]) # ordercheck
datamatrix=datamatrix[,sampleannotation$Type!="WT"]
sampleannotation=sampleannotation[sampleannotation$Type!="WT",]
sampleannotation$Type=factor(sampleannotation$Type)
print(dim(datamatrix))
#dev/debug
useparprior=TRUE
```
The covariate/batch balance;
```{r}
print(table(sampleannotation[ , c("Batch", "Type")])  )
```

@@

```{r reproduced_figA1, dev='png', fig.width=8, fig.height=15, fig.cap="fig. 1", tidy=FALSE, results='hold', fig.show='hold'}
variatonmeasure = apply(datamatrix, 1, FUN= function(x){var(x)})
clustermatrix = datamatrix[order(variatonmeasure, decreasing=TRUE),][1:2698,]
pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
         color = colorRampPalette(c("red", "black", "green"))(n = 299),
         annotation = sampleannotation[, c("Batch", "Type")], 
         cellheight=NA, cellwidth=13,fontsize=12,  drop_levels=TRUE, show_rownames=F,
         main=paste("Reproduced Fig A1.",sep=""),border_color=NA,treeheight_row=0)
rm(variatonmeasure, clustermatrix)
```
The Heatmap does not look exactly as Fig. A1 in the paper. This could be due to undocumented transformations preformed in Johnson et al., for example log-transformation or standarization. And there are different ways of calculating variation for a probe. What parameters that were used for the clustring could also be different. But the purpose of the figure was to show that a batch effect is present in the data and espeically for batch 3. This is also achived in the reproduced figure.

@@

Batch3 = datamatrix[,sampleannotation$Batch=="3"]
rm(mat)
```

@@

There is no description in how the 770 genes were selected so we use the same approach as in the reprodution of Figure A.1.
```{r reproduced_figA2, dev='png', fig.width=8, fig.height=15, fig.show='hold'}
variatonmeasure = apply(EB3, 1, FUN= function(x){var(x)})
clustermatrix = EB3[order(variatonmeasure, decreasing=TRUE),][1:770,]
pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
         color = colorRampPalette(c("red", "black", "green"))(n = 299),
         annotation = sampleannotation[, c("Batch", "Type", "Cell")],  
         cellheight=NA, cellwidth=13,fontsize=12,  drop_levels=TRUE, show_rownames=F,
         main=paste("Reproduced Fig A2.",sep=""),border_color=NA,treeheight_row=0)
rm(variatonmeasure, clustermatrix)
```
Again the heatmap is not exactly as in Johnson et al, but the batch clustring is broken and the cell clones with the same number seem to cluster.

@@

We start with the data before ComBat adjustment. Limma works best with log transformed and between array normalized values, so first we set negative values to 1 and filter out probes that has this in more than half the samples. To ease later comparisons the same filter will also be applied to the EB3 data set. Then the test is run with the batch included as a blocking factor.
```{r}
negativeprobesfilter =( rowSums(datamatrix>=1) >= (0.5*ncol(datamatrix)) ) & 
                      ( rowSums(EB3>=1) >= (0.5*ncol(EB3)) )
thismat = datamatrix[negativeprobesfilter, ]
thismat[thismat<1]=1
thismat=normalizeBetweenArrays(log2(thismat))
Type = as.factor(sampleannotation$Type)
Block = as.factor(sampleannotation$Batch)
design = model.matrix(~0+Type+Block)
fit = lmFit(thismat, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)	
fit2 = contrasts.fit(fit, cont.matrix)
datamatrix_limma_res = eBayes(fit2)
rm(thismat, Type, Block, design, fit, cont.matrix, fit2)
print(table(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05))
```

@@

Since limma was not the tool used in Johnson et al.s workflow, we will first need to create the corresponding analysis in Limma for the whole ComBat adjusted data set in order to make a fair comparison.
```{r}
thismat = EB3[negativeprobesfilter, ]
thismat[thismat<1]=1
thismat=normalizeBetweenArrays(log2(thismat))
Type = as.factor(sampleannotation$Type)
Block = as.factor(sampleannotation$Batch)
design = model.matrix(~0 + Type)
fit = lmFit(thismat, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
EB3_limma_res = eBayes(fit2)
rm(thismat, Type, Block, design, fit, cont.matrix, fit2)
print(table(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05))

```
The number of genes below the significance threshold is fairly similar to the t-test.

@@

Lets assume one aim of this study was to obtain a list of genes that differed between control and treatment with a significant cut-off of q<0.05. Then handling the batch effect with ComBat adjustment will yield a list of `r sum(qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05)` wheras a list of `r sum(qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05)` probes would have been closer to the truth. Also it is of interest to note that it is mostly the same probes that are found.
```{r, tidy=FALSE}
table( qvalue(datamatrix_limma_res$p.value[,1])$qvalue<0.05 ,
         qvalue(EB3_limma_res$p.value[,1])$qvalue<0.05,
       dnn=c("limma", "ComBat"))
```
And if they were to use the top 1000 probes in a gene set test they found many of the same genes regardless of the use of ComBat.
```{r, tidy=FALSE}
table( rank(datamatrix_limma_res$p.value[,1])  <=1000 & 
         rank(EB3_limma_res$p.value[,1])<=1000   )
```
Our conclusion is that for this study the error introduced by the use of ComBat would only have a modest effect on the final result.

Additional sanity checks
---------
To substantiate that the result from the use of ComBat is less trustworthy than the alternative analysis we provide a few additional sanity checks.

@@

fit = lmFit(EBrand, design)
fit2 = contrasts.fit(fit, cont.matrix)
EBrand_limma_pvalues = eBayes(fit2)$p.value[,1]

fit = lmFit(randommatrix, design)
fit2 = contrasts.fit(fit, cont.matrix)
randommatrix_limma_pvalues = eBayes(fit2)$p.value[,1]

Block = as.factor(sampleannotation$Batch)
design = model.matrix(~0+Type+Block)
fit = lmFit(randommatrix, design)
cont.matrix = makeContrasts ( contrasts="TypeR-TypeC", levels=design) 
fit2 = contrasts.fit(fit, cont.matrix)
fit2 = eBayes(fit2)
randommatrix_limma_batch_pvalues = fit2$p.value[,1]

```
Plotting the p-values
```{r, dev='svg', fig.width=8, fig.height=8}
hist(EBrand_limma_pvalues, border="blue",  main="P-values, Random numbers", breaks=100, xlab="p-value")
hist(randommatrix_limma_pvalues, border="black", add=T, breaks=100)
hist(randommatrix_limma_batch_pvalues, border="red", add=T, breaks=100)
legend("topright", legend=c("ComBat adjusted", "Limma adjusted", "No adjustment"), text.col=c("blue", "black", "red"))
```
The p-values for ComBat adjusted random numbers are slighlty enriched for low p-values. Indicating that the same enrichment seen for the real data is also false.

@@

References
-------------

```{r, echo=FALSE, tidy=FALSE,  warning=FALSE, comment="", prompt=FALSE}
#x = lapply(includelibs, FUN=citation)
for(x in lapply(includelibs, FUN=citation))
{
  print(x, style="textVersion")
  print("", quote = FALSE,include.rownames=FALSE)
}

```

```{r sessionInfo, echo=FALSE, comment=""}
sessionInfo()
```
